---
title: "npdb_work"
author: "Tiasia Saunders"
date: "`r Sys.Date()`"
output: html_document
---

## NPDB Data Analysis 

#### The NPDB is the National Practitioner Data Bank, which is a repository of reports on healthcare practitioners, including adverse actions and medical malpractice payments. 

##### Analysis Purpose: Through this analysis, I will explore the NPDB data to identify trends and patterns in adverse actions and medical malpractice reports across different states. 

##### Data Limitations: 
- For this analysis, I downloaded two datasets from the NPDB website: 
- The first dataset contains state-level data on adverse actions and medical malpractice reports from 2020- March 31st, 2025. 
- The second dataset contains yearly reports from 2015 to March 2025 of adverse actions and medical malpractice reports combined by state.


```{r}
library(tidyverse)
library(tidycensus)
library(dplyr)
library(janitor)
library(ggplot2)
```

```{r}
## loading data 
 npdb_state_data <- read_csv("npdb_state_record.csv") |> 
  clean_names()
 yearly_reports_data <- read_csv("yearly_reports.csv") |>
  clean_names()
```


##### Questions: 

1. What are the top 5 states with the most adverse action reports? 
2. What are the top 5 states with the most medical malpractice report? 
3. Do a timeline of the reports for the top states? 
4. What are the 5 top states with the least amount of reports as a whole (look at yearly_reports_data)? 



```{r}
top_adverse_states <- npdb_state_data |> 
  select(state, adverse_action_report) |> 
  arrange(desc(adverse_action_report)) |> 
  slice(2:6)
```

- The top 5 states that have the highest adverse action reports are the following: California, Texas, Florida, Ohio and Michigan 

## A bar chart of the top 5 states by Adverse Action Reports 
```{r}
top_adverse_states |> 
ggplot(aes(x = reorder(state, adverse_action_report), 
             y = adverse_action_report)) +
  geom_col(fill = "#2c3e50") +  
  coord_flip() +
  labs(
    title = "Top 5 States by Adverse Action Reports",
    x = "State",
    y = "Adverse Action Reports"
  ) +
  theme_minimal()
```


```{r}
top_malpractice_states <- npdb_state_data |> 
  select(state, medical_malpractice_payment_report) |> 
  arrange(desc(medical_malpractice_payment_report)) |> 
  slice(2:6)
```

### A bar chart of the top 5 states with Medical Malpractice Reports 
```{r}
top_malpractice_states |> 
ggplot(aes(x = reorder(state, medical_malpractice_payment_report), 
             y = medical_malpractice_payment_report)) +
  geom_col(fill = "lavender") +  
  coord_flip() +  
  labs(
    title = "Top 5 States by Medical Malpractice Reports",
    x = "State",
    y = "Medical Malpractice Reports"
  ) +
  theme_minimal()

```

- The top 5 states that have the highest medical malpractice payment reports are the following: New York, Florida, California, Pennsylvania and Texas 

- The states that are the occur on both lists are the following: California, Texas, and Florida  

### The top five states combining both type of reports 

```{r}
top_states_combined <- npdb_state_data |> 
  select(state, total) |> 
  arrange(desc(total)) |> 
  slice(2:6)

```

### A bar chart of the top 5 states combining both types of reports
```{r}
top_states_combined|> 
ggplot(aes(x = reorder(state,total), 
             y = total)) +
  geom_col(fill = "red") +  
  coord_flip() +  
  labs(
    title = "Top 5 States Combining Adverse and Malpractice Reports",
    x = "State",
    y = "Combined Reports"
  ) +
  theme_minimal()


```


```{r}
# Filter for California, Florida, and Texas
top_states <- yearly_reports_data |> 
  filter(state %in% c("California", "Florida", "Texas")) |> 
  select(-total)  # Remove 'Total' column

# Reshape from wide to long format
top_states <- top_states |> 
  pivot_longer(
    cols = starts_with("year_"),
    names_to = "Year",
    values_to = "Count"
  ) |> 
  mutate(
    Year = as.numeric(str_remove(Year, "year_"))  # Extract just the year number
  )
```

### A line chart of the trends in yearly counts for the top states
```{r}
ggplot(top_states, aes(x = Year, y = Count, color = state)) +
  geom_line(size = 1.2) +
  labs(
    title = "Trends in Yearly Counts for Selected States",
    x = "Year",
    y = "Count",
    color = "State"
  ) +
  theme_minimal()
```


- I plan to continue to do analysis on the NPDB data to identify trends and patterns in adverse actions and medical malpractice reports across different states. 
