---
title: "phmsa_work"
author: "Tiasia Saunders"
date: "2025-06-30"
output: html_document
---

# PHMSA Docket 2019-0091: Pipeline Safety: Amendments to Liquefied Natural Gas Facilities Comments Analysis 

### PHMSA’s Docket 2019-0091 is an advance notice of proposed rulemaking seeking public input on updating safety regulations for liquefied natural gas (LNG) facilities. The agency aims to modernize outdated rules, incorporate updated industry standards, and address emerging technologies and practices in LNG operations. This process includes evaluating environmental impacts and potential costs and benefits of regulatory changes.

```{r}
# Load libraries 
library(tidyverse)
library(tidytext)
library(dplyr)
library(ggplot2)
library(tm)
library(topicmodels)
library(janitor)
library(tm)
library(topicmodels)
library(lda)
library(ldatuning)
library(DT)
library(knitr)
library(kableExtra)
library(reshape2)
library(wordcloud)
library(pals)
library(SnowballC)
library(flextable)
library(rio)
library(spacyr)
spacy_install()

```

```{r}
# Load the data

read_csv("phmsa_comments_all.csv") |> 
  clean_names() -> phmsa_comments
```
## Leading States by Number of Comments
```{r}
state_comments <- phmsa_comments |> 
  mutate(state_province = ifelse(state_province %in% state.name, 
  state.abb[match(state_province, state.name)], state_province)) |> 
  select(state_province) |> 
  count(state_province, sort = TRUE) 
  print(state_comments)
```


- Top 5 states that commented on this docket were the following: CA, NY, FL, PA, and IL. This makes sense considering that CA, PA, and IL are part of the top states that has the most railroads. Additionally, PA and TX are a couple of states that produces the most LNG in the US. 

```{r}
time_comments <- phmsa_comments |> 
 select(posted_date, title) |> 
  group_by(posted_date) |> 
  summarize(comment_count = n())
  
```

## Graphic of Comments Over Time 
```{r}
ggplot(time_comments, aes(x = posted_date, y = comment_count)) +
  geom_line(size = 1) +
  labs(
    title = "PHMSA LNG Facilities Amendments Comments Over Time",
    x = "Posted Date",
    y = "Number of Comments", 
     caption = ".\nGraphic by Tiasia Saunders 07/08/2025"
  ) +
  theme_minimal()

```

- The docket was published on 05/05/25 and it started receiving comments 06/07 with one comment and started to spike on 06/09 with 2757 Comments, on 06/20 with 3627 comments, and on 06/30 with 297 comments.

```{r}
spiked_earlyjune_comments <- phmsa_comments |> 
  filter(as.character(posted_date) == "2025-06-09 04:00:00")

spiked_midjune_comments <- phmsa_comments |> 
  filter(as.character(posted_date) == "2025-06-20 04:00:00")

spiked_latejune_comments <- phmsa_comments |> 
  filter(as.character(posted_date) == "2025-06-30 04:00:00")
```

- So, the 2757 comments on June 06 came from Earthwork's submission box, so the organization must of posted a link to their submission box link on their social media to prompt the public to send the automated message in to the docket. 

- Same for the 3627 comments on June 20, include a different mass message, which I am attempting to find the submission box for it now. 

- The 297 comments on June 30, include mass comments from Earthworks and the other unknown submission box. 

## Comment Differantion 
```{r}
different_commentors <- phmsa_comments |> 
  select(title) |> 
  count(title, sort = TRUE) 
 
```

```{r}
# People: have both first and last names
people <- phmsa_comments |> 
  filter(!is.na(first_name) & first_name != "",
         !is.na(last_name) & last_name != "")

# Organizations: missing either first or last name
organizations <- phmsa_comments |> 
  filter(is.na(first_name) | first_name == "" |
         is.na(last_name) | last_name == "")
print(organizations)
```

#### Organization Submissions Results Analysis: 

- Organization Submissions came from multiple organizations including [Pipeline Safety Trust](https://pstrust.org/), an org dedicating to keep communities and environments safe from pipeline hazards and RESTORE, [Restore Explicit Symmetry to our Ravaged Earth](https://chorusfoundation.org/what-we-fund/restore-restore-explicit-symmetry-to-our-ravaged-earth-trust/), an organization that strive to raise public, governmental, and corporate awareness about the necessity to return to a sustainable equilibrium. 
- Pipeline Safety Trust's comment included an attachment in which they request a 60-day extension of the comment period for PHMSA's ANPRMS Docket 2019-0091 and Docket 2025-0019, citing the need for more time to review the complex issues, discuss with stakeholders, and provide meaningful input. They emphasize the importance of public safety and the need for thorough consideration of the proposed amendments to LNG facilities regulations.
- Other organizations include BLUE Engineering and Consulting, GPA Midstream Association, We Act for Environmental Justice, National Association of Pipeline Safety Representatives, Center for Biological Diversity Food & Water Watch, Science and Environmental Health Network, Micah Six Eight Mission, Cheniere Energy, Earthjustice, Environmental Defense Fund, National Grid, Eversource Energy, Sempra Infrastructure Partners, LP, South Texas Environmental Justice Network, American Petroleum Institute, Center for LNG, Interstate Natural Gas Association of America, AGA, APGA, NGA, The Williams Companies, Inc., Pipeline Safety Trust, Southwest Gas Corporation, CH·IV International, GTT North America
- It was 31 comments from these organizations 
- The organization commentators strongly emphasize the urgent need to revise outdated LNG safety regulations, citing unresolved risks and inadequate, secretive modeling that threatens public safety, especially in Southwest Louisiana’s LNG terminal zones. They highlight past workshops and expert recommendations that PHMSA has yet to fully address or make publicly available, including important test results from facilities like the LGFSTF. Concerns are raised about transparency, the influence of industry-aligned institutions, and the lack of meaningful community involvement in the regulatory process. Several comments also point to recent reports and real-world risks, including insurance companies withdrawing support due to underestimated dangers. Overall, they demand that PHMSA prioritize safety, transparency, and scientifically validated standards before proceeding with any rule revisions.



```{r}
same_earthworks_message <- phmsa_comments |> 
  filter(str_detect(comment, fixed("Dear Director Duffy,")))

same_other_message <- phmsa_comments |> 
  filter(str_detect(comment, fixed("I am writing,")))
  
```

#### Earthworks Same Message Submissions Analysis: 

- The same message came from Earthworks' submission [link](https://act.earthworks.org/page/83887/action/1?locale=en-US), urging PHMSA to update outdated LNG safety regulations to prioritize public safety, transparency, and accountability over industry profits. It called for strong methane leak standards and input from impacted communities to prevent pipeline failures and protect lives.

- The majority of the comments came from individuals, with 6,935 comments, with 2957 being the mass message from Earthwork's submission link. This suggests that the public is more concerned about the proposed amendments to LNG facilities than organizations are.

 
## Sentiment Analysis 

```{r}
#Tokenizing data 
phmsa_comments <- phmsa_comments %>% 
  mutate(clean_comment = str_replace_all(comment, "- ", ""))
tokenized_df <- tibble(
  comment = phmsa_comments$clean_comment
)
tokenized_comments <- phmsa_comments %>% 
  select(clean_comment) %>%
  unnest_tokens(word, clean_comment)

tokenized_comments


```

```{r}
# Removing stop words 
tokenized_comments <- tokenized_comments |> 
 anti_join(stop_words, by = c("word" = "word")) 
```

```{r}
# Creating Bigrams 
lng_bigrams <- tokenized_df |> 
  unnest_tokens(bigram, comment, token="ngrams", n=2)

lng_bigrams

#Filter out stop words.

lng_bigrams1 <- lng_bigrams |> 
  separate(bigram, c("word1", "word2"), sep = " ")

lng_filtered_bigrams <- lng_bigrams1 |> 
  filter(!word1 %in% stop_words$word) |> 
  filter(!word2 %in% stop_words$word) |> 
    filter(!is.na(word1)) |> 
     filter(!is.na(word2))

lng_counted_bigrams <- lng_filtered_bigrams |> 
  count(word1, word2, sort = TRUE)




```

```{r}
# Top twenty bigrams
top_twenty_bigrams <- lng_counted_bigrams |> 
  arrange(desc(n), word1, word2) |>
  head(20)

top_twenty_bigrams
```

```{r}
# Combine the bigram into one string for labeling
top_twenty_bigrams <- lng_counted_bigrams |>
  arrange(desc(n), word1, word2) |>
  head(20) |>
  mutate(bigrams = paste(word1, word2))

ggplot(top_twenty_bigrams, aes(x = reorder(bigrams, n), y = n)) +
  geom_col(fill = "pink") +
  coord_flip() +
  labs(
    title = "Top 20 Most Frequent Bigrams",
    x = "Bigrams",
    y = "Frequency", 
      caption = ".\nGraphic by Tiasia Saunders 07/08/2025"
  ) +
  theme_minimal()
```

- The top five most frequent bigrams are the following: lng export, protect workers, methane gas, lng infrastructure, and pipeline infrastructure. This suggests that the public is concerned about the safety and infrastructure of LNG facilities, as well as the impact on workers. The bigrams also suggest that the public is concerned about the environmental impact of LNG facilities, particularly in relation to methane gas emissions.
```{r}
nrc_sentiments <- get_sentiments("nrc")
afinn_sentiments <- get_sentiments("afinn")
```

```{r}
sentiments_comments <- tokenized_comments |> 
  inner_join(nrc_sentiments, by = "word") 
  overall_sentiment_count <- sentiments_comments |>
  count(sentiment) |> 
  arrange(desc(n))

```
```{r}
ggplot(overall_sentiment_count, aes(x = sentiment, y = n)) +
  geom_bar(stat = "identity", position = "dodge", fill = "lavender") +
  labs(
    title = "LNG Facilities Amendment Docket Comments Sentiment Count",
    x = "Sentiment",
    y = "Count", 
      caption = "\nGraphic by Tiasia Saunders 06/30/2025"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


- The most common sentiment of the comments is positive, suggesting that the public is generally supportive of the proposed amendments to LNG facilities. 
- The second most common sentiment is negative, indicating that there are some concerns about the proposed amendments. 
- The third most common sentiment is fear, suggesting that there are some fears about the potential impact of the proposed amendments on LNG facilities. 
- The fourth common sentiment is trust, indicating that there are some concerns about the trustworthiness of the proposed amendments. 


## Leading Positive, Negative, Fear, and Trust Sentiments 

```{r}
overall_negative_count <- sentiments_comments |> 
   filter(sentiment == "negative") |> 
  count(sentiment,word) |> 
  arrange(desc(n))

print(overall_negative_count)

```

```{r}
overall_positive_count <- sentiments_comments |> 
   filter(sentiment == "positive") |> 
  count(sentiment,word) |> 
  arrange(desc(n))

print(overall_positive_count)
```

```{r}
overall_fear_sentiment <- sentiments_comments |> 
  filter(sentiment == "fear") |> 
  count(sentiment,word) |> 
  arrange(desc(n)) 
  print(overall_fear_sentiment)

```

```{r}
overall_trust_sentiment <- sentiments_comments |> 
  filter(sentiment =="trust") |>
  count(sentiment,word) |> 
  arrange(desc(n)) 
print(overall_trust_sentiment)
```

## Sentiment Analysis Breakdown
```{r}
sentiments_comments %>%
  count(sentiment, word) %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>%  
  ungroup() %>%
  mutate(word = reorder_within(word, n, sentiment)) %>%  
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +  
  scale_y_reordered() + 
  labs(
    x = "Sentiment Analysis Breakdown For PHMSA Docket 2019-0091 Comments",
    y = NULL,
    caption = ".\nGraphic by Tiasia Saunders  07/01/2025"
  ) 
```

- The top five positive sentiments are: protect, effective, public, community, and strengthen which suggests that the public is generally supportive of the proposed amendments to LNG facilities. The public is concerned about the safety and effectiveness of LNG facilities, as well as the impact on the community. The public is also concerned about the strength of the proposed amendments and their potential impact on LNG facilities.
- The top five negative sentiments are: dangerous, leakage, hazardous, government, and reject, which suggests that the public is concerned about the potential dangers and hazards of LNG facilities. The public is also concerned about the government's role in regulating LNG facilities and the potential for LNG facilities to be rejected by the public.
- The top five fear sentiments are: dangerous, concerned, hazardous, government, and reject, which suggests that the public is concerned about the potential dangers and hazards of LNG facilities. The public is also concerned about the government's role in regulating LNG facilities and the potential for LNG facilities to be rejected by the public.
- The top five trust sentiments are: effective, strengthening, center, devastating, and safe, which suggests that the public is concerned about the overall effectiveness of the proposed LNG facilities amendments. Additionally, the public is concerned about the potential for LNG facilities to be strengthened and made safer. The public is also concerned about the potential for LNG facilities to be devastating to the environment and the community.


## Topic Modeling 

```{r}
textdata <- phmsa_comments |>  
  
  select(document_id, posted_date, title, first_name, last_name, comment) |> 
  as.data.frame() |> 
  rename(doc_id = document_id, text = comment)

# load stopwords
english_stopwords <- readLines("https://slcladal.github.io/resources/stopwords_en.txt", encoding = "UTF-8")
# create corpus object
corpus <- Corpus(DataframeSource(textdata))
# Preprocessing chain
processedCorpus <- tm_map(corpus, content_transformer(tolower))
processedCorpus <- tm_map(processedCorpus, removeWords, english_stopwords)
processedCorpus <- tm_map(processedCorpus, removePunctuation, preserve_intra_word_dashes = TRUE)
processedCorpus <- tm_map(processedCorpus, removeNumbers)
processedCorpus <- tm_map(processedCorpus, stemDocument, language = "en")
processedCorpus <- tm_map(processedCorpus, stripWhitespace)
```

```{r tm3a}
#DTM: rows correspond to the documents in the corpus. Columns correspond to the terms in the documents. Cells correspond to the weights of the terms. (Girder)
# compute document term matrix with terms >= minimumFrequency
minimumFrequency <- 5
DTM <- DocumentTermMatrix(processedCorpus, control = list(bounds = list(global = c(minimumFrequency, Inf))))
# have a look at the number of documents and terms in the matrix
dim(DTM)
# due to vocabulary pruning, we have empty rows in our DTM
# LDA does not like this. So we remove those docs from the
# DTM and the metadata
sel_idx <- slam::row_sums(DTM) > 0
DTM <- DTM[sel_idx, ]
textdata <- textdata[sel_idx, ]
#5 term minimum[1] 1387 3019
#5 term minimum[1] 308597 10339

```


```{r tm12}
# number of topics
# K <- 20
K <- 6
# set random number generator seed
set.seed(9161)
#Latent Dirichlet Allocation, LDA
topicModel2 <- LDA(DTM, K, method="Gibbs", control=list(iter = 500, verbose = 25, alpha = 0.2))
tmResult <- posterior(topicModel2)
theta <- tmResult$topics
beta <- tmResult$terms
topicNames <- apply(terms(topicModel2, 10), 2, paste, collapse = " ")  # reset topicnames
```

```{r}
# Step 1: Check dimensions
# Ensure doc_id and rownames are character
rownames(theta) <- as.character(rownames(theta))
textdata$doc_id <- as.character(textdata$doc_id)

# Get common document IDs
common_ids <- intersect(rownames(theta), textdata$doc_id)

# Filter both datasets to only include common documents
theta_aligned <- theta[rownames(theta) %in% common_ids, ]
textdata_filtered <- textdata[textdata$doc_id %in% common_ids, ]

# Reorder textdata_filtered to match theta_aligned
textdata_filtered <- textdata_filtered[match(rownames(theta_aligned), textdata_filtered$doc_id), ]

# Check that everything lines up
if (!all(rownames(theta_aligned) == textdata_filtered$doc_id)) {
  stop("The document IDs still do not match. Please check the data alignment.")
}

# Print dimensions
n_theta <- nrow(theta_aligned)
n_textdata_filtered <- nrow(textdata_filtered)
cat("Number of aligned theta rows: ", n_theta, "\n")
cat("Number of aligned textdata rows: ", n_textdata_filtered, "\n")

# Combine data
topic_data <- data.frame(theta_aligned, text = textdata_filtered$text)

# Aggregate topic proportions by dom_vs_intl
topic_proportion_per_category <- aggregate(. ~ text, data = topic_data, FUN = mean)

# Rename topic columns (assuming K and topicNames are defined)
colnames(topic_proportion_per_category)[2:(K+1)] <- topicNames

# Reshape for visualization
vizDataFrame <- melt(topic_proportion_per_category, id.vars = "text")

# Optionally filter out a decade if needed
# vizDataFrame <- subset(vizDataFrame, decade != 1960)
```


#### Examine topic names

```{r}
#enframe(): Converts a named list into a dataframe.
topics <- enframe(topicNames, name = "number", value = "text") %>% 
  unnest(cols = c(text)) 
  
topics
```

### Topic Keywords and Assigned Categories

| Topic # | Keywords (truncated)                                            | Suggested Category Name       |
|--------:|------------------------------------------------------------------|-------------------------------|
| 1       | pipeline, infrastructure, community, LNG, safety, update, gas   | `pipeline_safety_update`      |
| 2       | safety, LNG, community, protect, export, worker, hazard, PHMSA  | `lng_safety_and_labor`        |
| 3       | safety, LNG, administration, PHMSA, hazardous materials         | `hazardous_materials_admin`   |
| 4       | pipeline, infrastructure, methane, gas, safety                  | `methane_gas_infrastructure`  |
| 5       | LNG, export, worker, infrastructure, prioritization             | `lng_export_prioritization`   |
| 6       | terminal, rewrite, PHMSA, worker, hazard, infrastructure        | `phmsa_terminal_regulation`   |

```{r}
# Add categories based on topic keywords
vizDataFrame <- vizDataFrame |>
  mutate(category = case_when(
    str_detect(str_squish(variable), "pipelin infrastructur communiti safeti lng seek updat methan effect gas") ~ "pipeline_safety_update",
    str_detect(str_squish(variable), "lng safeti phmsa regul protect communiti industri standard pipelin requir") ~ "lng_safety_and_labor",
    str_detect(str_squish(variable), "administr safeti pipelin phmsa materi infrastructur hazard lng communiti worker") ~ "hazardous_materials_admin",
    str_detect(str_squish(variable), "pipelin infrastructur communiti safeti lng seek methan gas effect updat") ~ "methane_gas_infrastructure",
    str_detect(str_squish(variable), "safeti lng communiti export protect worker hazard infrastructur phmsa strengthen") ~ "lng_export_prioritization",
    str_detect(str_squish(variable), "safeti lng communiti export protect worker hazard phmsa infrastructur center",) ~ "phmsa_terminal_regulation",
    TRUE ~ "other"
  ))

```



```{r}
theta2 <- as.data.frame(theta)

topic1 <- theta2 %>% 
  rownames_to_column(var = "file") |> # putting the rownames into a new column called file
  mutate(file = str_remove(file, "^X"),  # Remove leading 'X'
         line = str_extract(file, "(?<=\\.txt)\\.\\d+")) |>   # Extract number after .txt
  mutate(file = str_remove(file, "\\.\\d+$")) |> 
  rename(topic1 = '1') |> # looking at first topic: pipelin infrastructur communiti lng safeti seek updat effect gas methan 
  top_n(20, topic1) |> 
  arrange(desc(topic1)) |>  
  select(file, line, topic1) 

```

```{r}
# Assuming `theta` is a dataframe related
theta2 <- as.data.frame(theta)

phmsa_topic <- theta2 %>%
  rename(phmsa_topic = '2') %>%  
  top_n(20, phmsa_topic) %>%
  arrange(desc(phmsa_topic)) %>% 
  select(phmsa_topic)

# Apply rownames_to_column to include story IDs
phmsa_topic <- tibble::rownames_to_column(phmsa_topic, "story_id")

# Clean up story_id if necessary (e.g., removing "X" or other unwanted characters)
phmsa_topic$story_id <- gsub("X", "", phmsa_topic$story_id)

# Check the top 20 story IDs
head(phmsa_topic$story_id, 20)
# Now it’s validated for PHMSA topics

```
```{r}
dominantTopics <- vizDataFrame |>
  group_by(text) |>
  slice_max(value, n = 1) |>   # Keep only the topic with highest probability
  ungroup()

```

```{r}

dominantTopics |>
  count(category) |>
  mutate(proportion = n / sum(n)) |>
  ggplot(aes(x = category, y = proportion, fill = category)) +
  geom_bar(stat = "identity") +
  ylab("Proportion of Comments") +
  labs(
    title = "Dominant Topics in PHMSA Comments",
    subtitle = "Based on most probable topic per comment",
    caption = "Graphic by Tiasia Saunders"
  ) +
  scale_fill_manual(
    values = c("#9933FF", "#33FFFF", "red", "yellow", "darkblue", "lavender")
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```



### Comments Topic Modeling Results: 

- So, the top two topics are: phmsa_terminal_regulation, lng_export_prioritization. This concludes that the majority of the comments are focused on safety regulations and prioritization of LNG exports. The public is worried about the safety of LNG facilities and the prioritization of LNG exports over safety concerns.

- The least mentioned topics are: hazardous_materials_admin, pipeline_safety_update, and methane_gas_infrastructure, and lng_safety_and_labor. This suggests that the public is less concerned about the administration of hazardous materials, updates to pipeline safety regulations, the infrastructure for methane gas, and the safety and labor practices. 

- These results are interesting because they show that the public is most concerned about safety regulations, labor issues, and prioritization of LNG exports, which are all important issues in the LNG industry. The least mentioned topics are less critical to the public, but they are still important for the public to address. 



