---
title: "kaggle"
author: "Tiasia Saunders"
date: "2025-07-21"
output: html_document
---


```{r}
# Load necessary libraries

library(tidyverse)
library(ggplot2)
library(dplyr)
library(janitor)
```

```{r}
# Load the dataset
netflix_tv_movies <- read_csv("netflix_tv_movies.csv")
mental_health <- read_csv("mental_health.csv")  |>
  clean_names()
```

How many movies? How many shows?

```{r}
netflix_tv_movies |> 
  count(type)

```
- There are 3407 movies and 1876 shows 


What are the top 5 movies with the most run times?
What year did the movie release?

```{r}
netflix_tv_movies |> 
  select(runtime, type, title, release_year) |>
  filter(type == "MOVIE") |> 
  arrange(desc(runtime)) |> 
  slice(1:5) 



```
- The top 5 movies with the longest runttimes are No Longer Kid, Once Upon a Time in America, A Lion in the House, Lagaan:Once Upon a Time in India, and Jodhaa Akbar. 

What is the breakdown of the age certification ratings for the entire dataset?
```{r}
netflix_tv_movies |> 
  count(age_certification)

```

What are the age certifications of the top 10 highest rated content?
What type of content receive the highest imdb scores?
```{r}
  netflix_tv_movies |> 
  select(age_certification, title, imdb_score) |>
  arrange(desc(imdb_score)) |> 
   slice(1:10)

```

Don't think we can answer the following questions considering there is no genre listed in the table and we can't do by year since it doesn't have the year that Netflix put it on their platform, instaed they just have the relased_year for when the movie came out.  

What genre dominates Netflix content by year?
Has the imdb rating for Netflix content reduced by year?



New Questions: 
- Make a bar chart of the top 10 imdb shows

```{r}
top_ten_movies <- netflix_tv_movies |> 
  select(title, type, imdb_score) |> 
  filter(type == "MOVIE") |> 
  arrange(desc(imdb_score)) |> 
  slice(1:10)
```


```{r}
ggplot(netflix_tv_movies, aes(x = title, y = imdb_score)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Domestic vs International Coverage Over Time",
    x = "Publication Date",
    y = "Number of Articles",
    color = "Outlet Type", 
    caption = ".\nGraphic by Tiasia Saunders (redacted - peer review)  05/11/2025"
  ) +
  theme_minimal()


```

- What are all the movies/shows released in 1993?
- Make a word chart of the top 10 imdb shows descriptions? 
- What are the top imdb shows with an R rating? 


## Mental Health stuff 

Basic Questions: 
1. What is the gender count in this dataset? 
2. What are the top 5 professions listed in this data? 
3. What is the breakdown of working student or professional? 
4. 

```{r}
gender_breakdown <- mental_health |> 
  count(gender)
print(gender_breakdown)

```
- There are 42,538 females and 51,262 males in this data 


```{r}
top_professions <- mental_health |> 
  count(profession) |> 
  arrange(desc(n)) |>
  slice(1:6)
print(top_professions)

```

- So, the top five professions are N/A, Teacher with 16,385, Content Writer with 5,187, Architect with 2,982, Consultant with 2920, and Pharmacist with 2656. 


```{r}
mental_health |> 
  count(working_professional_or_student)
```
- There are 18,772 students and 75,028 working professionals in this data. 

What are the dietary habits for students 

```{r}
mental_health |> 
  select(profession, dietary_habits) |> 
  filter(profession == "Student") 


```




