---
title: "asvab_work"
author: "Tiasia Saunders"
date: "2025-04-25"
output:
  html_document: default
  pdf_document: default
---
  
```{r}
# load libraries  
library(tidyverse)
library(readxl)
library(rio)
library(janitor)
library(ggplot2)
library(scales)

```

```{r}
# loading data 
asvab_data <- read_excel("asvab_data_2020-2024.xlsx")
```

```{r}
# clean heading names 
asvab_data <- asvab_data|> 
  clean_names()
```


```{r}

# clean data 
asvab_clean <- asvab_data %>%
  filter(if_any(everything(), ~ !is.na(.))) %>%
  mutate(across(where(is.character), str_trim)) %>%
  mutate(across(where(is.character), ~ na_if(., ""))) %>%
  mutate(across(where(is.character), ~ na_if(., "N/A"))) %>%
  mutate(across(where(is.character), ~ na_if(., "Unknown"))) 
  
```




## Summarizing the data 
  
  ### Race distribution 

```{r}
race_counts <- asvab_clean |> 
  separate_rows(race, sep = ",\\s*") |>   # splits by comma and optional space
  count(race, sort = TRUE) 

print(race_counts)
```

 ### Gender distribution 
```{r}
gender_count <- asvab_clean |> 
  count(sex, sort = TRUE) |> 
  print()

```

 ### Graphic of gender distribution 
```{r}
ggplot(gender_count, aes(x = sex, y = n, fill = sex)) +
  geom_col() +
  scale_fill_manual(values = c("Male" = "#3182bd", "Female" = "#e6550d")) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "ASVAB Gender Distribution",
       x = "Gender",
       y = "Count") +
  theme_minimal()

```
-- There are 332,846 males and 156518	females, meaning that about 32% percent is represented by females and about 56% percent is represented by males. 

  ### Age distribution 
  
```{r}
age_count <- asvab_clean |> 
  count(age, sort=TRUE) |> 
  print()
```
  ### Producing age ranges 
```{r}
asvab_clean <- asvab_clean |> 
  mutate(age_range = cut(
    age,
    breaks = c(17, 24, 34, 44, 54, 64, Inf),  # customize as needed
    labels = c("16–24", "25–34", "35–44", "45–54", "55–64", "65+"),
    right = TRUE,   # include the upper bound (e.g., 24 is part of 18–24)
    include.lowest = TRUE
  ))


```

```{r}
asvab_clean |> 
  count(age_range)
```

```{r}
asvab_clean$age_range <- factor(asvab_clean$age_range,
                                levels = c("16–24", "25–34", "35–44", "45–54", "55–64", "65+", "NA"))

# Plot
ggplot(asvab_clean, aes(x = age_range)) +
  geom_bar(fill = "lavender") +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_discrete(drop = FALSE) +  
  labs(title = "Age Range Distribution",
       x = "Age Range",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```
-- So mostly 16-24 year olds take the ASVAB test followed by the age range 25-34. 



### Latest Schooling By Race 
```{r}
latest_schooling_by_race <- asvab_clean |> 
  separate_rows(race, sep = ",\\s*") |>   
  group_by(race, latest_school_type) |> 
  summarise(count = n(), .groups = "drop") |> 
  arrange(desc(count))

```

### Graph of Latest Schooling By Race 
```{r}

ggplot(latest_schooling_by_race, aes(x = race, y = count, fill = latest_school_type)) +
  geom_col(position = "dodge") +
  labs(title = "Latest Schooling Type by Race",
       x = "Race",
       y = "Count",
       fill = "School Type") +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Latest Schooling By Gender 

```{r}
latest_schooling_by_gender <- asvab_clean |> 
    group_by(sex, latest_school_type) |> 
  summarise(count = n(), .groups = "drop")

print(latest_schooling_by_gender) 
  
```

### Graph of Latest Schooling By Gender 
```{r}
ggplot(latest_schooling_by_gender, aes(x = sex, y = count, fill = latest_school_type)) +
  geom_col(position = "dodge") +
  labs(title = "Latest Schooling Type by Gender",
       x = "Gender",
       y = "Count",
       fill = "School Type") +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


### Composite Scores Breakdown 

** The U.S. Air Force does not have a single, universal minimum composite score for any composite score. Instead, it has different minimum scores for various career fields. So, I will be using the min scores for different career fields that fall under each composite score section.  **

Breakdown of U.S. Air Force Composite Scores: 
- G	General: Verbal Expression (WK + PC) + Arithmetic Reasoning (AR)
- M	Mechanical: Mechanical Comprehension (MC) + Auto & Shop (AS)
- A	Administrative: Verbal Expression (WK + PC)
- E	Electronics: EI + AR + MK + GS

 G (General) - composite is used for general aptitude jobs that require verbal and math skills.

M (Mechanical) - focuses on mechanical and technical knowledge.

A (Administrative) - is for clerical and admin roles, based on verbal skills only.

E (Electronics) - combines several technical and science-based subtests for electronics and cyber-related roles.

Questions: 
** Which gender and which race is more associated with each composite score section? **
** What races and what genders qualify for what type of careers? ** 


### General Scores

```{r}

general_scores <- asvab_clean |> 
  select(asvab_ar, asvab_pc, asvab_wk, race, sex)
general_scores <- general_scores |>
    mutate(total_score = (asvab_ar + asvab_pc + asvab_wk))
print(general_scores)

```


### General Eligibility 
- Using 55 as the min score: qualify for roles such as Administrative (A) and Knowledge Operations (3D0X1)) 

```{r}
# Filter for individuals with score >= 55
eligible_general <- general_scores |>
  filter(total_score >= 55)

general_grouped_eligible <- eligible_general |>
  group_by(race, sex) |>
  summarize(
    count_eligible = as.numeric(n())) |> 
  arrange(desc(count_eligible))

print(general_grouped_eligible)

```

```{r}

# Plot: Eligible counts by race and sex with angled x-axis
g_plot <- ggplot(general_grouped_eligible, aes(x = race, y = count_eligible, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "General Eligibility Counts (Score ≥ 55) by Race and Sex",
    x = "Race",
    y = "Number Eligible",
    fill = "Sex", 
    capition = "Graphic by Tiasia Saunders"
  ) +
  scale_y_continuous(labels = scales::comma) +  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggsave("general_eligibility_counts.png", plot = g_plot, width = 12, height = 7, dpi = 300, bg = "white")

```

### Mechanical: Mechanical Comprehension (MC) + Auto & Shop (AS)
- Using 47 as the minimum Mechanical (M) composite score: qualifying for roles such as Aircraft Maintenance (2A5X1) and other general mechanical positions.

```{r}
mechanical_scores <- asvab_clean |> 
  select(asvab_as, asvab_mc,asvab_ar, race, sex)
mechanical_scores <- mechanical_scores |>
  mutate(total_score = (asvab_as +  asvab_mc + asvab_ar))
print(mechanical_scores)

```



```{r}
# Filter for individuals with score >= 47
eligible_mechanical <- mechanical_scores |>
  filter(total_score >= 47)

mechanical_grouped_eligible <- eligible_mechanical |>
  group_by(race, sex) |>
  summarize(
    count_eligible = as.numeric(n())) |> 
  arrange(desc(count_eligible))

print(mechanical_grouped_eligible)

```

```{r}

# Plot: Eligible counts by race and sex with angled x-axis
m_plot <- ggplot(mechanical_grouped_eligible, aes(x = race, y = count_eligible, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Mechanical Eligibility Counts (Score ≥ 47) by Race and Sex",
    x = "Race",
    y = "Number Eligible",
    fill = "Sex", 
    capition = "Graphic by Tiasia Saunders"
  ) +
  scale_y_continuous(labels = scales::comma) +  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("mechanical_eligibility_counts.png", plot = m_plot, width = 12, height = 7, dpi = 300, bg = "white")

```

## A	Administrative: Verbal Expression (WK + PC)
```{r}
admin_scores <- asvab_clean |> 
  select(asvab_wk, asvab_pc, race, sex)
admin_scores <- admin_scores |>
  mutate(total_score = (asvab_wk + asvab_pc))
print(admin_scores)

```


```{r}
# Filter for individuals with score >= 55
eligible_admin <- admin_scores |>
  filter(total_score >= 55)

admin_grouped_eligible <- eligible_admin |>
  group_by(race, sex) |>
  summarize(
    count_eligible = as.numeric(n()),  
  ) |>
  arrange(desc(count_eligible))

print(admin_grouped_eligible)

```

```{r}

# Plot: Eligible counts by race and sex with angled x-axis
a_plot <- ggplot(admin_grouped_eligible, aes(x = race, y = count_eligible, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Administrative Eligibility Counts (Score ≥ 55) by Race and Sex",
    x = "Race",
    y = "Number Eligible",
    fill = "Sex", 
    capition = "Graphic by Tiasia Saunders"
  ) +
  scale_y_continuous(labels = scales::comma) +  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("admin_eligibility_counts.png", plot = a_plot, width = 12, height = 7, dpi = 300, bg = "white")

```

## E	Electronics: EI + AR + MK + GS

```{r}
electronics_scores <- asvab_clean |> 
  select(asvab_ei, asvab_ar, asvab_mk, asvab_gs, race, sex)
electronics_scores <- electronics_scores |>
  mutate(total_score = (asvab_ar + asvab_mk + asvab_ei + asvab_gs))
print(electronics_scores)

```


```{r}
# Filter for individuals with score >= 60
eligible_electronics <- electronics_scores |>
  filter(total_score >= 60)

electronics_grouped_eligible <- eligible_electronics |>
  group_by(race, sex) |>
  summarize(
    count_eligible = as.numeric(n()),
    .groups = "drop"
  ) |> 

    arrange(desc(count_eligible))


print(electronics_grouped_eligible)


```


```{r}
# Plot: Eligible counts by race and sex with angled x-axis
 e_plot <- ggplot(electronics_grouped_eligible, aes(x = race, y = count_eligible, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Electronics Eligibility Counts (Score ≥ 60) by Race and Sex",
    x = "Race",
    y = "Number Eligible",
    fill = "Sex", 
    caption = "Graphic by Tiasia Saunders"
  ) +
  scale_y_continuous(labels = scales::comma) +  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Save to file - adjust width and height (in inches) and dpi as needed
ggsave("electronics_eligibility_counts.png", plot = e_plot, width = 12, height = 7, dpi = 300, bg = "white")

```

### Results: 

- The four graphics for the most part show similiar elgibility across all races with almost the same count of people per race that qualify for the positions. This makes sense considering the mininumn elgibility score is similiar ( 55 for General and Admin and 47 for Mechanical and 60 for Electronics)

Different ASVAB subtest sums → but similar high scores + similar cutoff thresholds → similar eligibility counts across composites.

If you can think of a better way to highlight the disparities in each composite scoring (in the process of researching how to do that) please let me know. But, for the most part for each composite scoring White Americans have high eligility count, followed by Black Americans, then Asian. 
`