---
title: "asvab_work"
author: "Tiasia Saunders"
date: "2025-04-25"
output:
  html_document: default
  pdf_document: default
---
## load libraries    
```{r}
library(tidyverse)
library(readxl)
library(rio)
library(janitor)
library(ggplot2)
library(scales)

```

## loading data 
```{r}
asvab_data <- read_excel("asvab_data_2020-2024.xlsx")
```


## clean heading names 
```{r}
asvab_data <- asvab_data|> 
  clean_names()
```

## clean data 

```{r}
asvab_clean <- asvab_data %>%
  # Remove rows with all NA values (optional)
  filter(if_any(everything(), ~ !is.na(.))) %>%
  
  # Trim whitespace from character columns
  mutate(across(where(is.character), str_trim)) %>%
  
  # Convert empty strings to NA
  mutate(across(where(is.character), ~ na_if(., ""))) %>%
  
  # Handle known inconsistent values (e.g., "N/A", "NA", "Unknown")
  mutate(across(where(is.character), ~ na_if(., "N/A"))) %>%
  mutate(across(where(is.character), ~ na_if(., "Unknown"))) 
  
```




## summarizing the data 
  
  ### race distribution 

```{r}
race_counts <- asvab_clean |> 
  separate_rows(race, sep = ",\\s*") |>   # splits by comma and optional space
  count(race, sort = TRUE) 

print(race_counts)
```

 ### gender distribution 
```{r}
gender_count <- asvab_clean |> 
  count(sex, sort = TRUE) |> 
  print()

```

 ### Graphic of gender distribution 
```{r}
ggplot(gender_count, aes(x = sex, y = n, fill = sex)) +
  geom_col() +
  scale_fill_manual(values = c("Male" = "#3182bd", "Female" = "#e6550d")) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "ASVAB Gender Distribution",
       x = "Gender",
       y = "Count") +
  theme_minimal()

```
-- there are 332,846 males and 156518	females

  ### age distribution 
  
```{r}
age_count <- asvab_clean |> 
  count(age, sort=TRUE) |> 
  print()
```
  ### producing age ranges 
```{r}
asvab_clean <- asvab_clean %>%
  mutate(age_range = cut(
    age,
    breaks = c(17, 24, 34, 44, 54, 64, Inf),  # customize as needed
    labels = c("16–24", "25–34", "35–44", "45–54", "55–64", "65+"),
    right = TRUE,   # include the upper bound (e.g., 24 is part of 18–24)
    include.lowest = TRUE
  ))


```

## age range 
```{r}
asvab_clean |> 
  count(age_range)
```

```{r}
asvab_clean$age_range <- factor(asvab_clean$age_range,
                                levels = c("16–24", "25–34", "35–44", "45–54", "55–64", "65+", "NA"))

# Plot
ggplot(asvab_clean, aes(x = age_range)) +
  geom_bar(fill = "lavender") +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_discrete(drop = FALSE) +  # Prevent ggplot from dropping unused levels
  labs(title = "Age Range Distribution",
       x = "Age Range",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```
-- So mostly 16-24 year olds take the ASVAB test 



### Latest Schooling By Race 
```{r}

# Separate multiracial combinations into individual rows
latest_schooling_by_race <- asvab_clean %>%
  separate_rows(race, sep = ",\\s*") %>%  # Split "A, B" into separate rows
  group_by(race, latest_school_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count))

```

### Graph of Latest Schooling By Race 
```{r}

ggplot(latest_schooling_by_race, aes(x = race, y = count, fill = latest_school_type)) +
  geom_col(position = "dodge") +
  labs(title = "Latest Schooling Type by Race",
       x = "Race",
       y = "Count",
       fill = "School Type") +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Latest Schooling By Gender 

```{r}
latest_schooling_by_gender <- asvab_clean |> 
    group_by(sex, latest_school_type) |> 
  summarise(count = n(), .groups = "drop")

print(latest_schooling_by_gender) 
  
```

### Graph of Latest Schooling By Gender 
```{r}
ggplot(latest_schooling_by_gender, aes(x = sex, y = count, fill = latest_school_type)) +
  geom_col(position = "dodge") +
  labs(title = "Latest Schooling Type by Gender",
       x = "Gender",
       y = "Count",
       fill = "School Type") +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




### General Science Asvab scores by race 
```{r}

general_science_scores_collapsed <- asvab_clean %>%
  separate_rows(race, sep = ",\\s*") %>%  # Split multi-race values into individual rows
  group_by(race, asvab_gs) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(race, asvab_gs)

print(general_science_scores_collapsed)


```






