---
title: "asvab_work"
author: "Tiasia Saunders"
date: "2025-04-25"
output:
  html_document: default
  pdf_document: default
---
## load libraries    
```{r}
library(tidyverse)
library(readxl)
library(rio)
library(janitor)
library(ggplot2)
library(scales)

```

## loading data 
```{r}
asvab_data <- read_excel("asvab_data_2020-2024.xlsx")
```


## clean heading names 
```{r}
asvab_data <- asvab_data|> 
  clean_names()
```

## clean data 

```{r}
asvab_clean <- asvab_data %>%
  # Remove rows with all NA values (optional)
  filter(if_any(everything(), ~ !is.na(.))) %>%
  
  # Trim whitespace from character columns
  mutate(across(where(is.character), str_trim)) %>%
  
  # Convert empty strings to NA
  mutate(across(where(is.character), ~ na_if(., ""))) %>%
  
  # Handle known inconsistent values (e.g., "N/A", "NA", "Unknown")
  mutate(across(where(is.character), ~ na_if(., "N/A"))) %>%
  mutate(across(where(is.character), ~ na_if(., "Unknown"))) 
  
```




## summarizing the data 
  
  ### race distribution 

```{r}
race_counts <- asvab_clean |> 
  separate_rows(race, sep = ",\\s*") |>   # splits by comma and optional space
  count(race, sort = TRUE) 

print(race_counts)
```

 ### gender distribution 
```{r}
gender_count <- asvab_clean |> 
  count(sex, sort = TRUE) |> 
  print()

```

 ### Graphic of gender distribution 
```{r}
ggplot(gender_count, aes(x = sex, y = n, fill = sex)) +
  geom_col() +
  scale_fill_manual(values = c("Male" = "#3182bd", "Female" = "#e6550d")) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "ASVAB Gender Distribution",
       x = "Gender",
       y = "Count") +
  theme_minimal()

```
-- there are 332,846 males and 156518	females

  ### age distribution 
  
```{r}
age_count <- asvab_clean |> 
  count(age, sort=TRUE) |> 
  print()
```
  ### producing age ranges 
```{r}
asvab_clean <- asvab_clean %>%
  mutate(age_range = cut(
    age,
    breaks = c(17, 24, 34, 44, 54, 64, Inf),  # customize as needed
    labels = c("16–24", "25–34", "35–44", "45–54", "55–64", "65+"),
    right = TRUE,   # include the upper bound (e.g., 24 is part of 18–24)
    include.lowest = TRUE
  ))


```

## age range 
```{r}
asvab_clean |> 
  count(age_range)
```

```{r}
asvab_clean$age_range <- factor(asvab_clean$age_range,
                                levels = c("16–24", "25–34", "35–44", "45–54", "55–64", "65+", "NA"))

# Plot
ggplot(asvab_clean, aes(x = age_range)) +
  geom_bar(fill = "lavender") +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_discrete(drop = FALSE) +  # Prevent ggplot from dropping unused levels
  labs(title = "Age Range Distribution",
       x = "Age Range",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```
-- So mostly 16-24 year olds take the ASVAB test 



### Latest Schooling By Race 
```{r}

# Separate multiracial combinations into individual rows
latest_schooling_by_race <- asvab_clean %>%
  separate_rows(race, sep = ",\\s*") %>%  # Split "A, B" into separate rows
  group_by(race, latest_school_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count))

```

### Graph of Latest Schooling By Race 
```{r}

ggplot(latest_schooling_by_race, aes(x = race, y = count, fill = latest_school_type)) +
  geom_col(position = "dodge") +
  labs(title = "Latest Schooling Type by Race",
       x = "Race",
       y = "Count",
       fill = "School Type") +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Latest Schooling By Gender 

```{r}
latest_schooling_by_gender <- asvab_clean |> 
    group_by(sex, latest_school_type) |> 
  summarise(count = n(), .groups = "drop")

print(latest_schooling_by_gender) 
  
```

### Graph of Latest Schooling By Gender 
```{r}
ggplot(latest_schooling_by_gender, aes(x = sex, y = count, fill = latest_school_type)) +
  geom_col(position = "dodge") +
  labs(title = "Latest Schooling Type by Gender",
       x = "Gender",
       y = "Count",
       fill = "School Type") +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




### General Science Asvab scores by race 
```{r}

general_science_scores_collapsed <- asvab_clean %>%
  separate_rows(race, sep = ",\\s*") %>%  # Split multi-race values into individual rows
  group_by(race, asvab_gs) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(race, asvab_gs)

print(general_science_scores_collapsed)


```



### clerical is a composite score of those four sections

Clerical (CL) – AR, MK, PC and WK

Combat (CO) – AS, PC, MC, WK

Electronics (EL) – AR, EI, GS, MK

Field Artillery (FA) – AR, MK, MC

General Maintenance (GM) – AS, EI, GS, MK

General Technical (GT) -AR, PC, WK

Mechanical Maintenance (MM) – AS, EI, MC

Operators and Food (OF) – AS, PC, MC, WK

Surveillance and Communications (SC) – AR, AS, MC, PC, WK

Skilled Technical (ST) – GS, MK, MC, PC, WK

Questions: 
** Which gender and which race is more associated with each composite score section? **
** What races and what genders qualify for what type of careers? ** 




### Clerical Scores

```{r}

clerical_scores <- asvab_clean |> 
  select(asvab_ar, asvab_mk, asvab_pc, asvab_wk, race, sex)
clerical_scores <- clerical_scores |>
    mutate(total_score = (asvab_ar + asvab_mk + asvab_pc + asvab_wk))
print(clerical_scores)

```


```{r}
grouped_clerical_scores <- clerical_scores |> 
  group_by(race, sex) |> 
  summarize(
    mean_total_score = mean(total_score, na.rm = TRUE),
    total_score_sum = sum(total_score, na.rm = TRUE),
    count = n(),
    sd_total_score = sd(total_score, na.rm = TRUE)
  )
print(grouped_clerical_scores) 
```
### Graphic of the Average ASVAB Clerical Scores by Race and Gender (need to redo to try to show elibility-- ask shelita about it)
```{r}

ggplot(clerical_scores, aes(x = race, y = total_score, fill = sex)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  labs(
    title = "Average ASVAB Clerical Scores by Race and Gender",
    x = "Race",
    y = "Average Total Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Combat Score- Combat (CO) – AS, PC, MC, WK

```{r}
combat_scores <- asvab_clean |> 
  select(asvab_as, asvab_pc, asvab_mc, asvab_wk, race, sex)
combat_scores <- combat_scores |>
  mutate(total_score = (asvab_as + asvab_pc + asvab_mc + asvab_wk))
print(combat_scores)

```


```{r}
grouped_combat_scores <- combat_scores |> 
  group_by(race, sex) |> 
  summarize(
    mean_total_score = mean(total_score, na.rm = TRUE),
    total_score_sum = sum(total_score, na.rm = TRUE),
    count = n(),
    sd_total_score = sd(total_score, na.rm = TRUE)
  )
print(grouped_combat_scores) 

```

### Electronics (EL) – AR, EI, GS, MK
```{r}
electronics_scores <- asvab_clean |> 
  select(asvab_ar, asvab_ei, asvab_gs, asvab_mk, race, sex)
electronics_scores <- electronics_scores |>
  mutate(total_score = (asvab_ar + asvab_ei + asvab_gs + asvab_mk))
print(electronics_scores)
```

### Field Artillery (FA) – AR, MK, MC
```{r}
field_artillery_scores <- asvab_clean |> 
  select(asvab_ar, asvab_mk, asvab_mc, race, sex) 
field_artillery_scores <- field_artillery_scores |>
  mutate(total_scores = (asvab_ar + asvab_mk + asvab_mc)) 
print(field_artillery_scores)
```

### General Maintenance (GM) – AS, EI, GS, MK

```{r}

general_maintence_scores<- asvab_clean |> 
  select( asvab_as, asvab_ei, asvab_gs, asvab_mk, race, sex )
general_maintence_scores <- general_maintence_scores |>
  mutate(total_score = (asvab_as + asvab_ei + asvab_gs + asvab_mk)) |> 
print(general_maintence_scores)
  

```

### General Technical (GT) -AR, PC, WK
```{r}
general_technical_scores <- asvab_clean |> 
  select(asvab_ar, asvab_pc, asvab_wk, race, sex ) 
general_technical_scores<- general_technical_scores |> 
  mutate(total_score = (asvab_ar + asvab_pc + asvab_wk)) |>
  print(general_technical_scores)

```

### Mechanical Maintenance (MM) – AS, EI, MC
```{r}
mechanical_maintence_scores <- asvab_clean |> 
  select( asvab_as, asvab_ei, asvab_mc, race, sex)
mechanical_maintence_scores <- mechanical_maintence_scores |>
  mutate(total_score = (asvab_as + asvab_ei + asvab_mc)) |> 
  print(mechanical_maintence_scores)
```


Operators and Food (OF) – AS, PC, MC, WK

```{r}
operators_food_scores <- asvab_clean |> 
  select(asvab_as, asvab_pc, asvab_mc, asvab_wk, race, sex)
operators_food_scores <- operators_food_scores |>
  mutate(total_score = (asvab_as + asvab_pc + asvab_mc + asvab_wk)) |> 
  print(operators_food_scores)

```

### Operators and Food Eligibility (using 85 as the min score because of an online search) 
```{r}
# Filter for total_score > 91
above_85 <- operators_food_scores |> 
  filter(total_score >= 85)

# Count number of each race above 91
of_race_counts <- above_85 |> 
  count(race)

# Plot the counts
ggplot(of_race_counts, aes(x = race, y = n, fill = race)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Individuals by Race with Total Score >= 85",
       x = "Race",
       y = "Count") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```


Surveillance and Communications (SC) – AR, AS, MC, PC, WK

```{r}
surveilance_comms_score <- asvab_clean |> 
  select(asvab_ar, asvab_as, asvab_mc, asvab_pc, asvab_wk, race, sex)
surveilance_comms_score <- surveilance_comms_score |>
  mutate(total_score = (asvab_ar + asvab_as + asvab_mc + asvab_pc + asvab_wk)) |> 
  print(surveilance_comms_score)
```


Skilled Technical (ST) – GS, MK, MC, PC, WK

```{r}
skilled_technical <- asvab_clean |> 
  select(asvab_gs, asvab_mk, asvab_mc, asvab_pc, asvab_wk, race, sex) 
skilled_technical <- skilled_technical |>
  mutate(total_score = (asvab_gs + asvab_mk + asvab_mc + asvab_pc + asvab_wk)) |> 
  print(skilled_technical)

```

### Skilled Technican Eligbility (using 91 as the min score because of an online search) 

```{r}
# Filter for total_score > 91
above_91 <- skilled_technical |> 
  filter(total_score >= 91)

# Count number of each race above 91
st_race_counts <- above_91 |> 
  count(race)

# Plot the counts
ggplot(st_race_counts, aes(x = race, y = n, fill = race)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Individuals by Race with Total Score > 91",
       x = "Race",
       y = "Count") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

